---
title: "BeeBDC vignette"

output:
   rmarkdown::html_vignette:
     code_folding: show
     Float_TOC: True
     TOC_depth: 2
     gallery: yes
     highlight: tango
     lightbox: yes
     self_contained: yes
     thumbnails: yes
     
vignette: >
  %\VignetteIndexEntry{BeeBDC vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
--- 

```{r libraryChunk, load-packages, include=FALSE}
  # markdown stuff
library(rgnparser)
library(magrittr)
library(knitr)
library(rmarkdown)
library(rmdformats)
library(prettydoc)
library(htmltools)
library(pkgdown)

  # Core packages
library(devtools)
library(BiocManager)
library(purrr)
library(here)
library(renv)
library(BeeBDC)
library(bdc)
library(CoordinateCleaner)
library(dplyr)
library(readr)
library(stringr)
library(lubridate)
library(tidyselect)
```

```{r secretRootPath, include=FALSE}
  # Use here::i_am to set the root directory
here::i_am("vignettes/BeeBDC_main.Rmd")
  # Set a root path that will work on the local machine and also on GitHub actions
RootPath <- paste0(dirname(getwd()), "/inst/extdata/WebDir")
  # Create the working directory in the RootPath if it doesn't exist already
if (!dir.exists(paste0(RootPath, "/Data_acquisition_workflow"))) {
    dir.create(paste0(RootPath, "/Data_acquisition_workflow"), recursive = TRUE)
}

```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(error = TRUE, 
                      eval = TRUE, 
                      tidy = TRUE, 
                      warning = FALSE,
                      root.dir = normalizePath(paste0(dirname(getwd()), "/inst/extdata/WebDir")))
```


# 0.0 Script preparation 
## 0.1 Working directory 
Choose the path to the root folder in which all other folders can be found
```{r falseRootPath, eval = FALSE}
RootPath <- paste0("/your/path/here")
```
```{r CreateRootPath, warning=FALSE}
  # Create the working directory in the RootPath if it doesn't exist already
if (!dir.exists(paste0(RootPath, "/Data_acquisition_workflow"))) {
    dir.create(paste0(RootPath, "/Data_acquisition_workflow"), recursive = TRUE)
}
  # Set the working directory
setwd(paste0(RootPath,"/Data_acquisition_workflow"))
```

For the first time that you run BeeBDC, and if you want to use the renv package to manage your 
packages, you can install renv...

            install.packages("renv")

 and then initialise renv the project.
 
            renv::init(project = paste0(RootPath,"/Data_acquisition_workflow")) 
  If you have already initialised a project, you can instead just activate it.
```{r activate}
renv::activate(project = paste0(RootPath,"/Data_acquisition_workflow"))
```

## 0.2 Install packages (if needed) 

You may need to install gdal on your computer. This can be done on a Mac by using Homebrew in the terminal and the command "brew install gdal". 

To start out, you will need to install BiocManager, devtools, **ComplexHeatmap**, and **rnaturalearthhires** to then install and fully use BeeBDC.
```{r installPackages, message=FALSE, warning=FALSE, results=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!require("devtools", quietly = TRUE))
    install.packages("devtools")

devtools::install_github("ropensci/rnaturalearthhires")
BiocManager::install("ComplexHeatmap")
```
   
   Now install BeeBDC.

```{r installBeeBDC, results=FALSE, message=FALSE}
devtools::install_github("https://github.com/jbdorey/BeeBDC.git", ref = "main", 
                        force = FALSE,
                        auth_token = "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")
```

  
  Snapshot the renv environment.
  
```{r snapshot}
renv::snapshot(project = paste0(RootPath,"/Data_acquisition_workflow"),
                 prompt = FALSE)
```
      
  Set up the directories used by BeeBDC. These directories include where the data, figures, reports, etc. will be saved. The RDoc needs to be a path RELATIVE to the RootPath; i.e., the file path from which the two diverge.
  
```{r dirMaker}
BeeBDC::dirMaker(
    RootPath = RootPath,
    RDoc = "vignettes/BeeBDC_main.Rmd") %>%
      # Add paths created by this function to the .GlobalEnv
    list2env(envir = .GlobalEnv)
```
      
## 0.3 Load packages 
      
Load all packages from the list specified above, with the addition of "rnaturalearthhires".

```{r lapply_library, results=FALSE}
lapply(c("ComplexHeatmap", "BeeBDC", "magrittr"), 
       library, character.only = TRUE)
```


***
        
# 1.0 Data merge 

<div class="alert alert-info">
  <strong> Attention:</strong> <br>
Although each line of code has been validated, in order to save time knitting the RMarkdown document the next section is display only. If you are not data merging (section 1.0) or preparing the data (section 2.0), feel free to skip to Section 3.0 Initial flags.

</div>

## 1.1 Download ALA data 

Download ALA data and create a new file in the DataPath to put those data into. You should also 
first make an account with ALA in order to download your data — <https://auth.ala.org.au/userdetails/registration/createAccount>
      
      BeeBDC::atlasDownloader(path = DataPath,
               userEmail = "your@email.edu.au",
               atlas = "ALA",
               ALA_taxon = "Apiformes")


##  1.2 Import and merge ALA, SCAN, iDigBio, and GBIF data 
      
Supply the path to where the data is, the save_type is either "csv_files" or "R_file".


      DataImp <- BeeBDC::repoMerge(path = DataPath, 
                      occ_paths = BeeBDC::repoFinder(path = DataPath),
                      save_type = "R_file")



If there is an error in finding a file, run `repoFinder()` by itself to troubleshoot. For example:

      		
      			#BeeBDC::repoFinder(path = DataPath)
      			#OUTPUT:
      			#$ALA_data
      			#[1] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/ALA_galah_path/galah_download_2022-09-15/data.csv"
      
      			#$GBIF_data
      			#[1] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/GBIF_webDL_30Aug2022/0000165-220831081235567/occurrence.txt"
      			#[2] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/GBIF_webDL_30Aug2022/0436695-210914110416597/occurrence.txt"
      			#[3] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/GBIF_webDL_30Aug2022/0436697-210914110416597/occurrence.txt"
      			#[4] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/GBIF_webDL_30Aug2022/0436704-210914110416597/occurrence.txt"
      			#[5] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/GBIF_webDL_30Aug2022/0436732-210914110416597/occurrence.txt"
      			#[6] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/GBIF_webDL_30Aug2022/0436733-210914110416597/occurrence.txt"
      			#[7] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/GBIF_webDL_30Aug2022/0436734-210914110416597/occurrence.txt"
      					
      			#$iDigBio_data
      			#[1] "F:/BeeDataCleaning2022/BeeDataCleaning/BeeDataCleaning/BeeData/iDigBio_webDL_30Aug2022/5aa5abe1-62e0-4d8c-bebf-4ac13bd9e56f/occurrence_raw.csv"
      
      			#$SCAN_data
      			#character(0)
      			#Failing because SCAN_data seems to be missing. Downloaded separatly from the one drive


Load in the most-recent version of these data if needed. This will return a list with: 
 
 1. The occurrence dataset with attributes and (.$Data_WebDL)
 2. The appended eml file (.$eml_files)


        DataImp <- BeeBDC::importOccurrences(path = DataPath,
                               fileName = "BeeData_")


## 1.3 Import USGS Data 

The `USGS_formatter()` will find, import, format and create metadata for the USGS dataset. The pubDate must be in day-month-year format.


      USGS_data <- BeeBDC::USGS_formatter(path = DataPath, pubDate = "19-11-2022")


## 1.4 Formatted Source Importer

Formatted source importer. Use this importer to find files that have been formatted and need to be added to the larger data file. 

The attributes file must contain "attribute" in its name, and the occurrence file must not.


      Complete_data <- BeeBDC::formattedCombiner(path = DataPath, 
                                    strings = c("USGS_[a-zA-Z_]+[0-9]{4}-[0-9]{2}-[0-9]{2}"), 
                                      # This should be the list-format with eml attached
                                    existingOccurrences = DataImp$Data_WebDL,
                                    existingEMLs = DataImp$eml_files) 


In the *catalogNumber*, remove ".*specimennumber:" as what comes after should be the USGS number to match for duplicates.


      Complete_data$Data_WebDL <- Complete_data$Data_WebDL %>%
        dplyr::mutate(catalogNumber = stringr::str_replace(catalogNumber,
                                                           pattern = ".*\\| specimennumber:",
                                                           replacement = ""))
      

## 1.5 Save data 

Choose the type of data format you want.


      BeeBDC::dataSaver(path = DataPath,# The main path to look for data in
           save_type = "CSV_file", # "R_file" OR "CSV_file"
           occurrences = Complete_data$Data_WebDL, # The existing datasheet
           eml_files = Complete_data$eml_files, # The existing EML files
           file_prefix = "Fin_") # The prefix for the fileNames
    rm(Complete_data, DataImp)


# 2.0 Data preperation 
The data preparatin section of the script relates mostly to integrating **bee** occurrence datasets and corrections and so may be skipped by many general taxon users.

## 2.1 standardise datasets

You may either use: 

- (a) the bdc import method (works well with general datasets) ***or*** 
- (b) the jbd import method (works well with above data merge)

### a. bdc import 

The bdc import is **NOT** truly supported here, but provided as an example. Please go to section 2.1b below.
Read in the bdc metadata and standardise the dataset to bdc.
      
            bdc_metadata <- readr::read_csv(paste(DataPath, "out_file", "bdc_integration.csv", sep = "/"))
            # ?issue — datasetName is a darwinCore field already!
            # Standardise the dataset to bdc
            db_standardized <- bdc::bdc_standardize_datasets(
              metadata = bdc_metadata,
              format = "csv",
              overwrite = TRUE,
              save_database = TRUE)
            # read in configuration description file of the column header info
            config_description <- readr::read_csv(paste(DataPath, "Output", "bdc_configDesc.csv",
                                                        sep = "/"), 
                                                  show_col_types = FALSE, trim_ws = TRUE)
      
### b. jbd import 

Find the path, read in the file, and add the *database_id* column.

      occPath <- BeeBDC::fileFinder(path = DataPath, fileName = "Fin_BeeData_combined_")


      db_standardized <- readr::read_csv(occPath, 
                                           # Use the basic ColTypeR function to determine types
                                         col_types = BeeBDC::ColTypeR(), trim_ws = TRUE) %>%
                                         dplyr::mutate(database_id = paste("Dorey_data_", 
                                         1:nrow(.), sep = ""),
                                         .before = family)


### c. optional thin


You can thin the dataset for ***TESTING ONLY!*** 

			 check_pf <- check_pf %>%
			   # take every 100th record
			   filter(row_number() %% 100 == 1)

***

## 2.2 Paige dataset

Paige Chesshire's cleaned American dataset — <https://doi.org/10.1111/ecog.06584>
 
### Import data
If you haven't figured out by now, don't worry about the column name warning — not all columns occur here.


      PaigeNAm <- readr::read_csv(paste(DataPath, "Paige_data", "NorAmer_highQual_only_ALLfamilies.csv",
                                        sep = "/"), col_types = BeeBDC::ColTypeR()) %>%
         # Change the column name from Source to dataSource to match the rest of the data.
        dplyr::rename(dataSource = Source) %>%
         # EXTRACT WAS HERE
          # add a NEW database_id column
        dplyr::mutate(
          database_id = paste0("Paige_data_", 1:nrow(.)),
          .before = scientificName)



<div class="alert alert-info">
  <strong> Attention:</strong> <br>
It is recommended to run the below code on the full bee dataset with more than 16GB RAM. Robert ran this on a laptop with 16GB RAM and an Intel(R) Core(TM) i7-8550U processor (4 cores and 8 threads) — it struggled.

</div>

### Merge Paige's data with downloaded data


      db_standardized <- BeeBDC::PaigeIntegrater(
          db_standardized = db_standardized,
          PaigeNAm = PaigeNAm,
            # This is a list of columns by which to match Paige's data to the most-recent download with. 
            # Each vector will be matched individually
          columnStrings = list(
            c("decimalLatitude", "decimalLongitude", 
              "recordNumber", "recordedBy", "individualCount", "samplingProtocol",
              "associatedTaxa", "sex", "catalogNumber", "institutionCode", "otherCatalogNumbers",
              "recordId", "occurrenceID", "collectionID"),         # Iteration 1
            c("catalogNumber", "institutionCode", "otherCatalogNumbers",
              "recordId", "occurrenceID", "collectionID"), # Iteration 2
            c("decimalLatitude", "decimalLongitude", 
              "recordedBy", "genus", "specificEpithet"),# Iteration 3
            c("id", "decimalLatitude", "decimalLongitude"),# Iteration 4
            c("recordedBy", "genus", "specificEpithet", "locality"), # Iteration 5
            c("recordedBy", "institutionCode", "genus", 
              "specificEpithet","locality"),# Iteration 6
            c("occurrenceID","decimalLatitude", "decimalLongitude"),# Iteration 7
            c("catalogNumber","decimalLatitude", "decimalLongitude"),# Iteration 8
            c("catalogNumber", "locality") # Iteration 9
          ) )


Remove spent data. 

      rm(PaigeNAm)



## 2.3 USGS
 The USGS dataset also partially occurs on GBIF from BISON. However, the occurrence codes are in a silly place... We will correct these here to help identify duplicates later.

        db_standardized <- db_standardized %>%
              # Remove the discoverlife html if it is from USGS
          dplyr::mutate(occurrenceID = dplyr::if_else(
            stringr::str_detect(occurrenceID, "USGS_DRO"),
            str_remove(occurrenceID, "http://www\\.discoverlife\\.org/mp/20l\\?id="),
            occurrenceID)) %>%
              # Use otherCatalogNumbers when occurrenceID is empty AND when USGS_DRO is detected there
          dplyr::mutate(
            occurrenceID = dplyr::if_else(
              stringr::str_detect(otherCatalogNumbers, "USGS_DRO") & is.na(occurrenceID),
              otherCatalogNumbers, occurrenceID)) %>%
               # Make sure that no eventIDs have snuck into the occurrenceID columns 
               # For USGS_DRO, codes with <6 digits are event ids
          dplyr::mutate(
            occurrenceID = dplyr::if_else(stringr::str_detect(occurrenceID, "USGS_DRO", negate = TRUE),
                 # Keep occurrenceID if it's NOT USGS_DRO
               occurrenceID, 
                 # If it IS USGS_DRO and it has => 6 numbers, keep it, else, NA
              dplyr::if_else(stringr::str_detect(occurrenceID, "USGS_DRO[0-9]{6,10}"),
                             occurrenceID, NA_character_)),
            catalogNumber = dplyr::if_else(stringr::str_detect(catalogNumber, "USGS_DRO", negate = TRUE),
                 # Keep catalogNumber if it's NOT USGS_DRO
              catalogNumber, 
                 # If it IS USGS_DRO and it has => 6 numbers, keep it, else, NA
              dplyr::if_else(stringr::str_detect(catalogNumber, "USGS_DRO[0-9]{6,10}"),
                             catalogNumber, NA_character_)))


## 2.4 Additional datasets
### Import additional and potentially private datasets.
**Note:** Private dataset functions are provided but the data itself is not integrated here until those datasets become freely available.

There will be some warnings were a few rows may not be formatted correctly or where dates fail to parse. This is normal.


###### a. EPEL
Guzman, L. M., Kelly, T. & Elle, E. A data set for pollinator diversity and their interactions with plants in the Pacific Northwest. Ecology, e3927 (2022). <https://doi.org/10.1002/ecy.3927>

    EPEL_Data <- BeeBDC::readr_BeeBDC(dataset = "EPEL",
                                    path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/bee_data_canada.csv",
                          outFile = "jbd_EPEL_data.csv",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
                          
###### b. Allan Smith-Pardo
Data from Allan Smith-Pardo 

    ASP_Data <- BeeBDC::readr_BeeBDC(dataset = "ASP",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/Allan_Smith-Pardo_Dorey_ready2.csv",
                          outFile = "jbd_ASP_data.csv",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
                          
###### c. Minckley
Data from Robert Minckley 

    BMin_Data <- BeeBDC::readr_BeeBDC(dataset = "BMin",
                                    path = paste0(DataPath, "/Additional_Datasets"),
                            inFile = "/InputDatasets/Bob_Minckley_6_1_22_ScanRecent-mod_Dorey.csv",
                            outFile = "jbd_BMin_data.csv",
                            dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
                            
###### d. BMont
Delphia, C. M. Bumble bees of Montana. <https://www.mtent.org/projects/Bumble_Bees/bombus_species.html>. (2022)

    BMont_Data <- BeeBDC::readr_BeeBDC(dataset = "BMont",
                                     path = paste0(DataPath, "/Additional_Datasets"),
                              inFile = "/InputDatasets/Bombus_Montana_dorey.csv",
                              outFile = "jbd_BMont_data.csv",
                              dataLicense = "https://creativecommons.org/licenses/by-sa/4.0/")
                              
###### e. Ecd
Ecdysis. Ecdysis: a portal for live-data arthropod collections, <https://serv.biokic.asu.edu/ecdysis/index.php> (2022).

    Ecd_Data <- BeeBDC::readr_BeeBDC(dataset = "Ecd",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/Ecdysis_occs.csv",
                          outFile = "jbd_Ecd_data.csv",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
                          
###### f. Gai
Gaiarsa, M. P., Kremen, C. & Ponisio, L. C. Pollinator interaction flexibility across scales affects patch colonization and occupancy. Nature Ecology & Evolution 5, 787-793 (2021). <https://doi.org/10.1038/s41559-021-01434-y>

    Gai_Data <- BeeBDC::readr_BeeBDC(dataset = "Gai",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/upload_to_scan_Gaiarsa et al_Dorey.csv",
                          outFile = "jbd_Gai_data.csv",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
                          
###### g. CAES

From the Connecticut Agricultural Experiment Station

    CAES_Data <- BeeBDC::readr_BeeBDC(dataset = "CAES",
                                    path = paste0(DataPath, "/Additional_Datasets"),
                            inFile = "/InputDatasets/CT_BEE_DATA_FROM_PBI.xlsx",
                            outFile = "jbd_CT_Data.csv",
                            sheet = "Sheet1",
                            dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")

###### h. GeoL

    GeoL_Data <- BeeBDC::readr_BeeBDC(dataset = "GeoL",
                                    path = paste0(DataPath, "/Additional_Datasets"),
                            inFile = "/InputDatasets/Geolocate and BELS_certain and accurate.xlsx",
                            outFile = "jbd_GeoL_Data.csv",
                            dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")


###### i. EaCO

    EaCO_Data <- BeeBDC::readr_BeeBDC(dataset = "EaCO",
                                    path = paste0(DataPath, "/Additional_Datasets"),
                            inFile = "/InputDatasets/Eastern Colorado bee 2017 sampling.xlsx",
                            outFile = "jbd_EaCo_Data.csv",
                            dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
    

###### j. FSCA

Florida State Collection of Arthropods

    FSCA_Data <- BeeBDC::readr_BeeBDC(dataset = "FSCA",
                                    path = paste0(DataPath, "/Additional_Datasets"),
                            inFile = "InputDatasets/fsca_9_15_22_occurrences.csv",
                            outFile = "jbd_FSCA_Data.csv",
                            dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")

###### k. Texas SMC
Published or unpublished data from Texas literature not in an online database, usually copied into spreadsheet from document format, or otherwise copied from a very differently-formatted spreadsheet. Unpublished or partially published data were obtained with express permission from the lead author.

    SMC_Data <- BeeBDC::readr_BeeBDC(dataset = "SMC",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/TXbeeLitOccs_31Oct22.csv", 
                          outFile = "jbd_SMC_Data.csv",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
                          

###### l. Texas Bal
Data with GPS coordinates from Ballare, K. M., Neff, J. L., Ruppel, R. & Jha, S. Multi-scalar drivers of biodiversity: local management mediates wild bee community response to regional urbanization. Ecol. Appl. 29, e01869 (2019), <https://doi.org/10.1002/eap.1869>. The version on Dryad is missing site GPS coordinates (by accident). Kim is okay with these data being made public as long as her paper is referenced. - Elinor Lichtenberg

    Bal_Data <- BeeBDC::readr_BeeBDC(dataset = "Bal",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/Beedata_ballare.xlsx", 
                          outFile = "jbd_Bal_Data.csv",
                          sheet = "animal_data",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")

###### m. Palouse Lic
*Attached: My canola data. I tried to get this in DarwinCore format. These data go with a manuscript currently under review. Here’s the bioRxiv version: Lichtenberg, E. M., Milosavljević, I., Campbell, A. J. & Crowder, D. W. Differential effects of soil conservation practices on arthropods and crop yield. bioRxiv, https://doi.org/10.1101/2021.1112.1106.471474 (2022) <https://www.biorxiv.org/content/10.1101/2021.12.06.471474v2>. These are the data I will be putting on SCAN. - Elinor Lichtenberg
 
    Lic_Data <- BeeBDC::readr_BeeBDC(dataset = "Lic",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/Lichtenberg_canola_records.csv", 
                          outFile = "jbd_Lic_Data.csv",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")

###### n. Arm

    Arm_Data <- BeeBDC::readr_BeeBDC(dataset = "Arm",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/Bee database Armando_Final.xlsx",
                          outFile = "jbd_Arm_Data.csv",
                          sheet = "Sheet1",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")

###### o. Dor
From several papers: <https://doi.org/10.3897/jhr.79.57308>, <https://doi.org/10.3897/jhr.81.59365>, and <https://doi.org/10.11646/zootaxa.4674.1.1>.

    Dor_Data <- BeeBDC::readr_BeeBDC(dataset = "Dor",
                                   path = paste0(DataPath, "/Additional_Datasets"),
                          inFile = "/InputDatasets/DoreyData.csv",
                          outFile = "jbd_Dor_Data.csv",
                          dataLicense = "https://creativecommons.org/licenses/by-nc-sa/4.0/")
    

### 2.5 Merge all 
Remove these spent datasets.

      rm(EPEL_Data, ASP_Data, BMin_Data, BMont_Data, Ecd_Data, Gai_Data, CAES_Data, 
      GeoL_Data, EaCO_Data, FSCA_Data, SMC_Data, Bal_Data, Lic_Data, Arm_Data, Dor_Data)
   
Read in and merge all. There are more `readr_BeeBDC()` supported than currently implemented and these represent datasets that will be publicly released in the future. See '?`readr_BeeBDC()`' for details.

      
    db_standardized <- db_standardized %>%
      dplyr::bind_rows(
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_ASP_data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_EPEL_data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_BMin_data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_BMont_data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_Ecd_data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_Gai_data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_CT_Data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_GeoL_Data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_EaCo_Data.csv"), col_types = BeeBDC::ColTypeR()), 
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_SMC_Data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_Bal_Data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_Lic_Data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_Arm_Data.csv"), col_types = BeeBDC::ColTypeR()),
        readr::read_csv(paste0(DataPath, "/Additional_Datasets", 
                               "/jbd_Dor_Data.csv"), col_types = BeeBDC::ColTypeR())) %>% 
        # END bind_rows
      suppressWarnings(classes = "warning") # End suppressWarnings — due to col_types


### 2.6 Match database_id 

If you have prior runs from which you'd like to match database_ids with from the current run. You may use the below script to try to match database IDs with prior runs. 


Read in a prior run of choice.


      priorRun <- BeeBDC::fileFinder(path = DataPath,
                              file = "01_prefilter_database_9Aug22.csv") %>%
        readr::read_csv(file = ., col_types = BeeBDC::ColTypeR())


This function will attempt to find the database_ids from prior runs.

      
      db_standardized <- BeeBDC::idMatchR(
      currentData = db_standardized,
      priorData = priorRun,
        # First matches will be given preference over later ones
      matchBy = tibble::lst(c("gbifID", "dataSource"),
                            c("catalogNumber", "institutionCode", "dataSource", "decimalLatitude",
                              "decimalLongitude"),
                            c("occurrenceID", "dataSource","decimalLatitude","decimalLongitude"),
                            c("recordId", "dataSource","decimalLatitude","decimalLongitude"),
                            c("id", "dataSource","decimalLatitude","decimalLongitude"),
                            # Because INHS was entered as it's own dataset but is now included in the GBIF    download...
                            c("catalogNumber", "institutionCode", "dataSource",
                              "decimalLatitude","decimalLongitude")),
        # You can exclude datasets from prior by matching their prefixs — before first underscore:
      excludeDataset = c("ASP", "BMin", "BMont", "CAES", "EaCO", "Ecd", "EcoS",
                         "Gai", "KP", "EPEL", "CAES", "EaCO", "FSCA", "SMC", "Lic", "Arm"))
    
     # Remove redundant files
    rm(priorRun)

Save the dataset.


      db_standardized %>%
        readr::write_excel_csv(.,
                         paste(OutPath_Intermediate, "00_prefilter_database.csv",
                               sep = "/"))


# 3.0 Initial flags 
Read data back in if needed. OutPath_Intermediate (and a few other directories) should be have been created and saved to the global environment by `dirMaker()`.


    if(!exists("db_standardized")){
      db_standardized <- readr::read_csv(paste(OutPath_Intermediate, "00_prefilter_database.csv",
                                        sep = "/"), col_types = BeeBDC::ColTypeR())}


<div class="alert alert-info">
  <strong> Attention:</strong> <br>
For the purposes of this website, we will only use a combination of two test datasets, that are provided with BeeBDC. These datasets are (i) 105 occurrence records across three species (**bees3sp**) and (ii) 100 records across 85 names (**beesRaw**). 

</div>

```{r 3.0}
data("bees3sp", package = "BeeBDC")
data("beesRaw", package = "BeeBDC")
db_standardized <- dplyr::bind_rows(beesRaw, 
                                      # Only keep a subset of columns from bees3sp
                             bees3sp %>% dplyr::select(tidyselect::all_of(colnames(beesRaw)), countryCode))
```




*For more details about the bdc package, please see their  [tutorial.](https://brunobrr.github.io/bdc/articles/prefilter.html)*

 
## 3.1 SciName 
Flag occurrences without *scientificName* provided.

```{r 3.1}
check_pf <- bdc::bdc_scientificName_empty(
  data = db_standardized,
  sci_name = "scientificName")
  # now that this is saved, remove it to save space in memory
rm(db_standardized) 
```


## 3.2 MissCoords 
Flag occurrences with missing *decimalLatitude* and *decimalLongitude*.

```{r 3.2}
check_pf <- bdc::bdc_coordinates_empty(
  data = check_pf,
  lat = "decimalLatitude",
  lon = "decimalLongitude")
```

## 3.3 OutOfRange 
Flag occurrences that are not on earth (outside of -180 to 180 or -90 to 90 degrees).

```{r 3.3}
check_pf <- bdc::bdc_coordinates_outOfRange(
  data = check_pf,
  lat = "decimalLatitude",
  lon = "decimalLongitude")
```

## 3.4 Source 
Flag occurrences that don't match the *basisOfRecord* types below.

```{r 3.4}
check_pf <- bdc::bdc_basisOfRecords_notStandard(
  data = check_pf,
  basisOfRecord = "basisOfRecord",
  names_to_keep = c(
    # Keep all plus some at the bottom.
    "Event",
    "HUMAN_OBSERVATION",
    "HumanObservation",
    "LIVING_SPECIMEN",
    "LivingSpecimen",
    "MACHINE_OBSERVATION",
    "MachineObservation",
    "MATERIAL_SAMPLE",
    "O",
    "Occurrence",
    "MaterialSample",
    "OBSERVATION",
    "Preserved Specimen",
    "PRESERVED_SPECIMEN",
    "preservedspecimen Specimen",
    "Preservedspecimen",
    "PreservedSpecimen",
    "preservedspecimen",
    "S",
    "Specimen",
    "Taxon",
    "UNKNOWN",
    "",
    NA,
    "NA",
    "LITERATURE", 
    "None", "Pinned Specimen", "Voucher reared", "Emerged specimen"
  ))
```

	   
## 3.5 CountryName 

Try to harmonise country names.  

### a. prepare dataset 

Fix up country names based on common problems above and extract ISO2 codes for occurrences.


```{r 3.5a}
check_pf_noNa <- BeeBDC::countryNameCleanR(
  data = check_pf,
    # Create a Tibble of common issues in country names and their replacements
  commonProblems = dplyr::tibble(problem = c('U.S.A.', 'US','USA','usa','UNITED STATES',
                                              'United States','U.S.A','MX','CA','Bras.','Braz.',
                                              'Brasil','CNMI','USA TERRITORY: PUERTO RICO'),
                                  fix = c('United States of America','United States of America',
                                          'United States of America','United States of America',
                                          'United States of America','United States of America',
                                          'United States of America','Mexico','Canada','Brazil',
                                          'Brazil','Brazil','Northern Mariana Islands','PUERTO.RICO'))
  )
```


### b. run function
Get country name from coordinates using a wrapper around the `jbd_country_from_coordinates()` function. Because our dataset is much larger than those used to design bdc, we have made it so that you can analyse data in smaller pieces. Additionally, like some other functions in BeeBDC, we have implemented parallel operations (using mc.cores = #cores in stepSize = #rowsPerOperation); see '?`jbd_CfC_chunker()`' for details.

```{r 3.5b, message=FALSE, warning=FALSE}
suppressWarnings(
  countryOutput <- BeeBDC::jbd_CfC_chunker(data = check_pf_noNa,
                                   lat = "decimalLatitude",
                                   lon = "decimalLongitude",
                                   country = "country",
                                    # How many rows to process at a time
                                   stepSize = 1000000,
                                    # Start row
                                   chunkStart = 1,
                                   path = OutPath_Intermediate,
                                   append = FALSE,
                                   mc.cores = 1),
  classes = "warning")
```

### c. re-merge 

Join these datasets.

```{r 3.5ci}
check_pf <- dplyr::left_join(check_pf, 
                             countryOutput, 
                             by = "database_id",
                             suffix = c("", "CO"))  %>% 
    # Take the new country name if the original is NA
  dplyr::mutate(country = dplyr::if_else(is.na(country),
                                         countryCO,
                                         country)) %>%
    # Remove duplicates if they arose from left_join!
  dplyr::distinct()
```

Save the dataset.

```{r 3.5cii}
check_pf %>%
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "01_prefilter_database.csv",
                         sep = "/"))
```

Read in if needed. 

```{r 3.5ciii}
if(!exists("check_pf")){
check_pf <- readr::read_csv(paste(DataPath, 
             "Output", "Intermediate", "01_prefilter_database.csv", sep = "/"),
             col_types = BeeBDC::ColTypeR())}
```

remove these interim datasets.

```{r 3.5civ}
rm(check_pf_noNa, countryOutput)
```

## 3.6 StandardCoNames 

Run the function (standardises country names and adds ISO2 codes if needed).

```{r 3.6}
  # Standardise country names and add ISO2 codes if needed
check_pf <- bdc::bdc_country_standardized(
  # Remove the countryCode and country_suggested columns to avoid an error with 
    # where two "countryCode" and "country_suggested" columns exist (i.e. if the dataset has been  
    # run before)
  data = check_pf %>% dplyr::select(!tidyselect::any_of(c("countryCode", "country_suggested"))),
  country = "country"
) 
```

## 3.7 TranspCoords 

Flag and correct records when *decimalLatitude* and *decimalLongitude* appear to be transposed. We created this chunked version of  `bdc::bdc_coordinates_transposed()` because it is very RAM-heavy using our large bee dataset. However, like many of our other 'jbd_...' functions there are other improvements - e.g., parallel running. 

```{r 3.7, message=FALSE, warning=FALSE}
check_pf <- BeeBDC::jbd_Ctrans_chunker(
  # bdc_coordinates_transposed inputs
  data = check_pf,
  id = "database_id",
  lat = "decimalLatitude",
  lon = "decimalLongitude",
  country = "country",
  countryCode = "countryCode",
  border_buffer = 0.2, # in decimal degrees (~22 km at the equator)
  save_outputs = TRUE,
  sci_names = "scientificName",
  # chunker inputs
  stepSize = 1000000,  # How many rows to process at a time
  chunkStart = 1,  # Start row
  append = FALSE,  # If FALSE it may overwrite existing dataset
  path = OutPath_Check,
  mc.cores = 1
) 
```

Get a quick summary of the number of transposed records.

```{r 3.7ii, eval = FALSE}
table(check_pf$coordinates_transposed, useNA = "always")
```

Save the dataset
```{r 3.7iii}
check_pf %>%
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "01_prefilter_database.csv",
                         sep = "/"))
gc()
```


Read data in again if needed 

```{r 3.7iv}
if(!exists("check_pf")){
  check_pf <- readr::read_csv(paste(OutPath_Intermediate, "01_prefilter_database.csv",
                                    sep = "/"))}
```


## 3.8 Coord-country 
Collect all country names in the *country_suggested* column. We rebuilt a bdc function to flag occurrences where the coordinates are inconsistent with the provided country name.

```{r 3.8}
check_pf <- BeeBDC::jbd_coordCountryInconsistent(
  data = check_pf,
  lon = "decimalLongitude",
  lat = "decimalLatitude",
  mapResolution = 50,
  pointBuffer = 0.01)
```

Save the dataset.
```{r 3.8ii}
check_pf %>%
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "01_prefilter_database.csv",
                         sep = "/"))
```

## 3.9 GeoRefIssue 

This function Identifies records whose coordinates can potentially be extracted from locality information (must be manually checked later).

```{r 3.9, eval = FALSE}
xyFromLocality <- bdc::bdc_coordinates_from_locality(
  data = check_pf,
  locality = "locality",
  lon = "decimalLongitude",
  lat = "decimalLatitude",
  save_outputs = TRUE
) %>%
# Save data 
  readr::write_excel_csv(paste(OutPath_Check, "01_coordinates_from_locality.csv",
                         sep = "/"))
```


Remove spent data.
```{r 3.9ii, eval = FALSE}
rm(xyFromLocality)
```

## 3.10 Flag Absent 
Flag the records marked as "absent".

```{r 3.10}
check_pf <- BeeBDC::flagAbsent(data = check_pf,
                   PresAbs = "occurrenceStatus")
```

## 3.11 flag License 

Flag the records that may not be used according to their license information.

```{r 3.11}
check_pf <- BeeBDC::flagLicense(data = check_pf,
                    strings_to_restrict = "all",
                    # DON'T flag if in the following dataSource(s)
                    excludeDataSource = NULL)
```

## 3.12 GBIF issue 

Flag select issues that are flagged by GBIF.

```{r 3.12}
check_pf <- BeeBDC::GBIFissues(data = check_pf, 
                   issueColumn = "issue", 
                   GBIFflags = c("COORDINATE_INVALID", "ZERO_COORDINATE")) 
```

## 3.13 Flag Reports 
### a. Save flags 

Save the flags so far. This function will make sure that you keep a copy of everything that has been flagged up until now. This will be updated throughout the script and can accessed at the end, so be wary of moving files around manually. However, these data will also still be maintained in the main running file and so this is an optional fail-safe..

```{r 3.13a}
flagFile <- BeeBDC::flagRecorder(
  data = check_pf,
  outPath = paste(OutPath_Report, sep =""),
  fileName = paste0("flagsRecorded_", Sys.Date(),  ".csv"),
    # These are the columns that will be kept along with the flags
  idColumns = c("database_id", "id", "catalogNumber", "occurrenceID", "dataSource"),
    # TRUE if you want to find a file from a previous part of the script to append to
  append = FALSE)

```

Update the *.summary* column

```{r 3.13b}
check_pf <- BeeBDC::summaryFun(
  data = check_pf,
    # Don't filter these columns (or NULL)
  dontFilterThese = NULL,
    # Remove the filtering columns?
  removeFilterColumns = FALSE,
    # Filter to ONLY cleaned data?
  filterClean = FALSE)
```



### c. Reporting 
Use bdc to generate reports.

```{r 3.13c, eval = FALSE}
(report <- bdc::bdc_create_report(data = check_pf,
                                  database_id = "database_id",
                                  workflow_step = "prefilter",
                                  save_report = TRUE)
)
```

## 3.14 Save 

Save the intermediate dataset.

```{r 3.14}
check_pf %>%
  readr::write_excel_csv(., paste(OutPath_Intermediate, "01_prefilter_output.csv",
                            sep = "/"))
```



# 4.0 Taxonomy 
*For more information about the corresponding bdc functions used in this section, see their [tutorial](https://brunobrr.github.io/bdc/articles/taxonomy.html). *

Read in the filtered dataset or rename the 3.x dataset for 4.0.

```{r 4.0}
if(!exists("check_pf")){
database <-
  readr::read_csv( paste(OutPath_Intermediate, "01_prefilter_output.csv",
                         sep = "/"), col_types = BeeBDC::ColTypeR())
}else{
    # OR rename and remove
  database <- check_pf
  # Remove spent dataset
  rm(check_pf)}
```

Remove *names_clean* if it already exists (i.e. you have run the following functions before on this dataset before).

```{r 4.0ii}
			database <- database %>%
			  dplyr::select(!tidyselect::any_of("names_clean"))
```

## 4.1 Prep data names 

This step cleans the database's *scientificName* column. 
	
***! MAC: You might need to install gnparser through terminal — brew
			brew tap gnames/gn
			brew install gnparser
			
			
<div class="alert alert-info">
  <strong> Attention:</strong> <br>
This can be difficult for a Windows install. Ensure you have the most recent version of R, Rstudio, and R packages. Also check package 'rgnparser' is installed correctly. If you still can not get the below code to work, you may have to download the latest version of 'gnparser' from [here](https://github.com/gnames/gnparser/releases/tag/v1.6.9). You may then need to manually install it and edit your systems environmental variable PATH to locate 'gnparser.exe'. See [here](https://github.com/gnames/gnparser#installation).

</div>			


```{r 4.1, eval = TRUE}
parse_names <-
  bdc::bdc_clean_names(sci_names = database$scientificName, save_outputs = FALSE)
```

Keep only the *.uncer_terms* and *names_clean* columns.

```{r 4.1ii}
parse_names <-
  parse_names %>%
  dplyr::select(.uncer_terms, names_clean)
```

Merge names with the complete dataset.
```{r 4.1iii}
database <- dplyr::bind_cols(database, parse_names)
rm(parse_names)
```

## 4.2 Taxo harmonization 
Read in the custom taxonomy file from the BeeBDC package and [Discover Life](http://discoverlife.org) website.

```{r 4.2}
data("beesTaxonomy", package = "BeeBDC")
```


Harmonise the names in the occurrence tibble. This flags the occurrences without a matched name and matches names to their correct name according to Discover Life. You can also use multiple cores to achieve this. See '?`harmoniseR()`' for details.

```{r 4.2ii}
database <- BeeBDC::harmoniseR(path = DataPath, #The path to a folder that the output can be saved
                       taxonomy = beesTaxonomy, # The formatted taxonomy file
                       data = database,
                       mc.cores = 4)
```


You don't need this file any more...
```{r 4.2iii}
rm(beesTaxonomy)
```



Save the harmonised file.

```{r 4.2iv}
database %>%
  readr::write_excel_csv(.,
                   paste(DataPath, "Output", "Intermediate", "02_taxonomy_database.csv",
                         sep = "/"))
```

## 4.3 Save flags 
SAVE the flags so far. This will find the most-recent flag file and append your new data to it. You can double-check the data and number of columns if you'd like to be thorough and sure that all of data are intact. <3 

```{r 4.3}
flagFile <- BeeBDC::flagRecorder(
  data = database,
  outPath = paste(OutPath_Report, sep =""),
  fileName = paste0("flagsRecorded_", Sys.Date(),  ".csv"),
  idColumns = c("database_id", "id", "catalogNumber", "occurrenceID", "dataSource"),
  append = TRUE,
  printSummary = TRUE)
```

# 5.0 Space

Read in the last database.

```{r 5.0}
if(!exists("database")){
database <-
  readr::read_csv(paste(OutPath_Intermediate, "02_taxonomy_database.csv", sep = "/"),
                  col_types = BeeBDC::ColTypeR())}
```

## 5.1 Coord precision 
This function identifies records with a coordinate precision below a specified number of decimal places. For example, the precision of a coordinate with 1 decimal place is 11.132 km at the equator, i.e., the scale of a large city. The major difference between the bdc and BeeBDC functions is that `jbd_coordinates_precision()` will only flag occurrences if BOTH lat and lon are rounded (as opposed to only one of these).
"Coordinates with one, two, or three decimal places present a precision of ~11.1 km, ~1.1 km, and ~111 m at the equator, respectively."

```{r 5.1}
check_space <-
  BeeBDC::jbd_coordinates_precision(
    data = database,
    lon = "decimalLongitude",
    lat = "decimalLatitude",
    ndec = 2 # number of decimals to be tested
  )
```

Remove the spent dataset.
```{r 5.1ii}
rm(database)
```

Save the harmonised file.
```{r 5.1iii}
check_space %>%
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "03_space_inter_database.csv",
                         sep = "/"))
```

## 5.2 Common spatial issues 
Only run for occurrences through clean_coordinates that are spatially 'valid'. 

```{r 5.2}
tempSpace <- check_space %>% 
  dplyr::filter(!.coordinates_empty == FALSE) %>%
  dplyr::filter(!.coordinates_outOfRange == FALSE)
```

Next, we will flag common spatial issues using functions of the package CoordinateCleaner. It addresses some common issues in biodiversity datasets.

```{r 5.2ii, message=TRUE, warning=FALSE}
tempSpace <-
  CoordinateCleaner::clean_coordinates(
    x =  tempSpace,
    lon = "decimalLongitude",
    lat = "decimalLatitude",
    species = "scientificName",
    countries = NULL, # Tests if coords are from x country. This is not needed.
    tests = c(
      "capitals",     # records within 0.5 km of capitals centroids
      "centroids",    # records within 1 km around country and province centroids
      "equal",      # records with equal coordinates
      "gbif",         # records within 1 km of GBIF headquarters. (says 1 degree in package, but code says 1000 m)
      "institutions", # records within 100m of zoo and herbaria
      "zeros"       # records with coordinates 0,0
      # "seas"        # Not flagged as this should be flagged by coordinate country inconsistent
    ),
    capitals_rad = 1000,
    centroids_rad = 500,
    centroids_detail = "both", # test both country and province centroids
    inst_rad = 100, # remove zoo and herbaria within 100m
    range_rad = 0,
    zeros_rad = 0.5,
    capitals_ref = NULL,
    centroids_ref = NULL,
    country_ref = NULL,
    country_refcol = "countryCode",
    inst_ref = NULL,
    range_ref = NULL,
    # seas_scale = 50,
    value = "spatialvalid" # result of tests are appended in separate columns
  ) %>%
  dplyr::select(!tidyselect::starts_with(".summary")) %>%
  dplyr::tibble()
```

Re-merge the datasets.

```{r 5.2iii}
check_space <- tempSpace %>%
  # Re-bind with the records that were removed earlier
  dplyr::bind_rows(check_space %>% 
                     dplyr::filter(.coordinates_empty == FALSE | 
                                     .coordinates_outOfRange == FALSE) ) 
```

Remove the temporary dataset.
```{r 5.2iv}
rm(tempSpace)
```

Save the intermediate dataset.

```{r 5.2v}
check_space %>%
  readr::write_excel_csv(paste(OutPath_Intermediate, "03_space_inter_database.csv",
                         sep = "/"))
```

## 5.3 Diagonal + grid 
Finds sequential numbers that could be fill-down errors in lat and long, and groups by the 'groupingColumns'. This is accomplished by using a sliding window with the length determined by minRepeats. Only coordinates of precision 'ndec' (number of decimals in decimal degree format) will be examined. Note, that this function is very RAM-intensive and so the use of multiple threads should be approached with caution depending on your dataset. However, the option is provided.

```{r 5.3}
check_space <- BeeBDC::diagonAlley(
  data = check_space,
  # The minimum number of repeats needed to find a sequence in for flagging
  minRepeats = 6,
  ndec = 3,
  groupingColumns = c("eventDate", "recordedBy", "datasetName"),
  mc.cores = 1)
```

Spatial gridding from rasterisation: Select only the records with more than X occurrences.

```{r 5.3ii}
griddingDF <- check_space %>%
  # Exclude NA lat and lon values
  tidyr::drop_na(c("decimalLatitude", "decimalLongitude")) %>%
  # Group by the dataset name
  dplyr::group_by(datasetName) %>%
  # Remove rows that aren't unique for lat and long
  dplyr::distinct(decimalLongitude, decimalLatitude,
                  .keep_all = TRUE) %>%
  # Find the groups with 4 or more occurrence records 
  dplyr::filter(dplyr::n() >= 4) %>%
  dplyr::ungroup()

```

Run the gridding analysis to find datasets that might be gridded.

```{r 5.3iii}
gridded_datasets <- CoordinateCleaner::cd_round(
  x = griddingDF,
  lon = "decimalLongitude",
  lat = "decimalLatitude",
  ds = "datasetName",
  T1 = 7,
  min_unique_ds_size = 4,
  test = "both",
  value = "dataset",
  graphs = FALSE,
  verbose = TRUE,
  reg_out_thresh = 2,
  reg_dist_min = 0.1,
  reg_dist_max = 2
) %>% 
  dplyr::tibble()
# The griddingDF is no longer needed. remove it.
rm(griddingDF)
```

Integrate these results with the main dataset.

```{r 5.3iv}
check_space <- check_space %>%
  # Join the datasets
  dplyr::left_join(
    # Select the columns of interest
    dplyr::select(gridded_datasets, dataset, lon.flag, lat.flag, summary),
    by = c("datasetName" = "dataset")) %>%
  # Make new columns with more-consistent naming and change the NA vlaues to = TRUE (not flagged)
  dplyr::mutate(.lonFlag = tidyr::replace_na(lon.flag, TRUE),
                .latFlag = tidyr::replace_na(lat.flag, TRUE),
                .gridSummary = tidyr::replace_na(summary, TRUE)) %>%
  # Remove old columns
  dplyr::select(!c(lon.flag, lat.flag, summary))
```

Save the gridded_datasets file for later examination.

```{r 5.3 v}
gridded_datasets %>%
  readr::write_excel_csv(paste(OutPath_Intermediate, "03_space_griddedDatasets.csv",
                         sep = "/"))
```

Now remove this file.
```{r 5.3vi}
rm(gridded_datasets)
```

## 5.4 Uncertainty 
Flag records that exceed a *coordinateUncertaintyInMeters* threshold.

```{r 5.4}
check_space <- coordUncerFlagR(data = check_space,
                               uncerColumn = "coordinateUncertaintyInMeters",
                               threshold = 1000)
```
## 5.5 Country Checklist 
Read in the country-level checklist.

```{r 5.5}
data("beesChecklist", package = "BeeBDC")
```


```{r 5.5ii}
check_space <- BeeBDC::countryOutlieRs(checklist = beesChecklist,
                        data = check_space,
                        keepAdjacentCountry = TRUE,
                        pointBuffer = 0.05,
                          # Scale of map to return, one of 110, 50, 10 OR 'small', 'medium', 'large'
                          # Smaller numbers will result in much longer calculation times. 
                          # We have not attempted a scale of 10.
                        rnearthScale = 50,
                        mc.cores = 4)
  # A list of failed species-country combinations and their numbers can be output here
check_space %>%
  dplyr::filter(.countryOutlier == FALSE) %>%
  dplyr::select(database_id, scientificName, country) %>%
  dplyr::group_by(scientificName) %>% 
  dplyr::mutate(count_scientificName = n()) %>%
  dplyr::distinct(scientificName, country, .keep_all = TRUE) %>% 
  readr::write_excel_csv(paste(OutPath_Intermediate, "03_space_failedCountryChecklist.csv",
                         sep = "/"))
```


## 5.6 Map spatial errors 

Rebuild the *.summary* column.

```{r 5.6}
check_space <- BeeBDC::summaryFun(
  data = check_space,
  dontFilterThese = NULL,
  removeFilterColumns = FALSE,
  filterClean = FALSE)
```


Use col_to_map in order to map ONE spatial flag at a time or map the *.summary* column for all flags. 

```{r 5.6ii}
check_space %>%
  dplyr::filter(.summary == FALSE) %>% # map only records flagged as FALSE
  bdc::bdc_quickmap(
    data = .,
    lon = "decimalLongitude",
    lat = "decimalLatitude",
    col_to_map = ".summary",
    size = 0.9
  )

```


## 5.7 Space report 

Create the space report using bdc.
```{r 5.7, eval = FALSE}
(report <-
   bdc::bdc_create_report(
     data = dplyr::tibble(check_space %>% dplyr::select(!.uncer_terms)),
     database_id = "database_id",
     workflow_step = "space",
     save_report = TRUE)
)
```

## 5.8 Space figures 

Create figures for the spatial data filtering results.

```{r 5.8}
(figures <-
    BeeBDC::jbd_create_figures(
      data = dplyr::tibble(check_space %>% dplyr::select(!.uncer_terms)),
      path = DataPath,
      database_id = "database_id",
      workflow_step = "space",
      save_figures = TRUE)
)
```


Check figure, the options are:
- *.cap* = Records around country capital centroid	
- *.cen* = Records around country or province centroids
- *.dbl* = Duplicated coordinates per species
- *.equ* = Identical coordinates
- *.otl* = Geographical outliers
- *.gbf* = Records around the GBIF headquarters
- *.inst* = Records around biodiversity institutions
- *.rou* = Rounded (probably imprecise) coordinates
- *.urb* = Records within urban areas — (Likely not relevant for bees.) 

			You can examine these figures, for example, by running:

			figures$.rou

Save interim dataset.

```{r 5.8ii}
check_space %>%
  readr::write_excel_csv(paste(OutPath_Intermediate, "03_space_inter_database.csv",
                         sep = "/"))
```

## 5.9 Save flags

Save the flags so far.
```{r 5.9}
BeeBDC::flagRecorder(
  data = check_space,
  outPath = paste(OutPath_Report, sep =""),
  fileName = paste0("flagsRecorded_", Sys.Date(),  ".csv"),
  idColumns = c("database_id", "id", "catalogNumber", "occurrenceID", "dataSource"),
  append = TRUE,
  printSummary = TRUE)
```

## 5.10 Save 

Save the intermediate dataset
```{r 5.10}
check_space %>%
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "03_space_database.csv",
                         sep = "/"))
```


# 6.0 Time 
Read in the last database, if needed.

```{r 6.0}
if(!exists("check_space")){
  check_time <-
    readr::read_csv(paste(OutPath_Intermediate, "03_space_database.csv", sep = "/"),
                    col_types = BeeBDC::ColTypeR())
  }else{
  check_time <- check_space
      # Remove the spent file
  rm(check_space)}
```


You can plot a histogram of dates here.

```{r 6.0ii}
hist(lubridate::ymd_hms(check_time$eventDate, truncated = 5), breaks = 20)
```

Filter some silly dates that don't make sense.

```{r 6.0iii}
check_time$year <- ifelse(check_time$year > lubridate::year(Sys.Date()) | check_time$year < 1600,
                        NA, check_time$year)
check_time$month <- ifelse(check_time$month > 12 | check_time$month < 1,
                         NA, check_time$month)
check_time$day <- ifelse(check_time$day > 31 | check_time$day < 1,
                       NA, check_time$day)
```
## 6.1 Recover dates

The `dateFindR()` function will search through some other columns in order to find and rescue dates that may not have made it into the correct columns. It will further update the *eventDate*, *day*, *month*, and *year* columns where these data were a) missing and b) located in one of the searched columns.

```{r 6.1}
check_time <- BeeBDC::dateFindR(data = check_time,
                        # Years above this are removed (from the recovered dates only)
                        maxYear = lubridate::year(Sys.Date()),
                        # Years below this are removed (from the recovered dates only)
                        minYear = 1700)
```


## 6.2 No eventDate 
Flag records that simply lack collection date. :*(

```{r 6.2}
check_time <-
  bdc::bdc_eventDate_empty(data = check_time, eventDate = "eventDate")
```


## 6.3 Old records 
This will flag records prior to the date selected. 1970 is frequently chosen for SDM work. You may not need to filter old records at all, so think critically about your use. We have chosen 1950 as a lower extreme.

```{r 6.3}
check_time <-
  bdc::bdc_year_outOfRange(data = check_time,
                           eventDate = "year",
                           year_threshold = 1950)
```

## 6.4 Time report 
*Not all of time, just the time pertaining to our precise occurrence records.*
Firstly, update the *.summary* column.

```{r 6.4}
check_time <- BeeBDC::summaryFun(
  data = check_time,
  # Don't filter these columns (or NULL)
  dontFilterThese = c(".gridSummary", ".lonFlag", ".latFlag", ".uncer_terms"),
  # Remove the filtering columns?
  removeFilterColumns = FALSE,
  # Filter to ONLY cleaned data?
  filterClean = FALSE)

( report <-
    bdc::bdc_create_report(data = check_time,
                           database_id = "database_id",
                           workflow_step = "time",
                           save_report = FALSE)
)   
```

## 6.5 Time figures 

Create time results figures.

```{r 6.5}
figures <-
  BeeBDC::jbd_create_figures(data = check_time,
                     path = DataPath,
                     database_id = "database_id",
                     workflow_step = "time",
                     save_figures = TRUE)
```

You can check figures by using...
```{r 6.5ii}
figures$year
```

Save the ~raw time dataset into the intermediate folder.

```{r 6.5iii}
check_time %>%
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "04_time_database.csv",
                         sep = "/"))
```


## 6.6 Save flags 
Save the flags so far.

```{r}
BeeBDC::flagRecorder(
  data = check_time,
  outPath = paste(OutPath_Report, sep =""),
  fileName = paste0("flagsRecorded_", Sys.Date(),  ".csv"),
  idColumns = c("database_id", "id", "catalogNumber", "occurrenceID", "dataSource"),
  append = TRUE,
  printSummary = TRUE)

```

# 7.0 De-duplication  
The raw dataset can be re-read here if it does not already exist.
```{r 7.0}
if(!exists("check_time")){
  check_time <-
    readr::read_csv(paste(OutPath_Intermediate, "04_time_database.csv",
                          sep = "/"),
                    col_types = BeeBDC::ColTypeR())}
```

## 7.1 deDuplicate
We will FLAG duplicates here. These input columns can be hacked to de-duplicate as you wish. This function uses user-specified inputs and columns to identify duplicate occurrence records. Duplicates are identified iteratively and will be tallied up, duplicate pairs clustered, and sorted at the end of the function. The function is designed to work with Darwin Core data with a *database_id* column, but it is also modifiable to work with other columns.
I would encourage you to, see '?`dupeSummary()`' for more details as this function is quite modifiable ot user needs.

```{r 7.1}
check_time <- BeeBDC::dupeSummary(
  data = check_time,
  path = OutPath_Report,
   # options are "ID","collectionInfo", or "both"
  duplicatedBy = "collectionInfo", 
    # The columns to generate completeness info from (and to sort by completness)
  completeness_cols = c("decimalLatitude",  "decimalLongitude",
                        "scientificName", "eventDate"),
   # The columns to ADDITIONALLY consider when finding duplicates in collectionInfo
  collectionCols = c("decimalLatitude", "decimalLongitude", "scientificName", "eventDate", 
                     "recordedBy"),
    # The columns to combine, one-by-one with the collectionCols
  collectInfoColumns = c("catalogNumber", "otherCatalogNumbers"),
    # Custom comparisons — as a list of columns to compare
     # RAW custom comparisons do not use the character and number thresholds
  CustomComparisonsRAW = dplyr::lst(c("catalogNumber", "institutionCode", "scientificName")),
     # Other custom comparisons use the character and number thresholds
  CustomComparisons = dplyr::lst(c("gbifID", "scientificName"),
                                  c("occurrenceID", "scientificName"),
                                  c("recordId", "scientificName"),
                                  c("id", "scientificName")),
   # The order in which you want to KEEP duplicated based on data source
   # try unique(check_time$dataSource)
  sourceOrder = c("CAES", "Gai", "Ecd","BMont", "BMin", "EPEL", "ASP", "KP", "EcoS", "EaCO",
                  "FSCA", "Bal", "SMC", "Lic", "Arm",
                  "USGS", "ALA", "GBIF","SCAN","iDigBio"),
    # Paige ordering is done using the database_id prefix, not the dataSource prefix.
  prefixOrder = c("Paige", "Dorey"),
    # Set the complexity threshold for id letter and number length
     # minimum number of characters when WITH the numberThreshold
  characterThreshold = 2,
     # minimum number of numbers when WITH the characterThreshold
  numberThreshold = 3,
     # Minimum number of numbers WITHOUT any characters
  numberOnlyThreshold = 5
) %>% # END dupeSummary
  dplyr::as_tibble(col_types = BeeBDC::ColTypeR())
```

Save the dataset into the intermediate folder.
```{r 7.1ii}
check_time %>%
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "04_2_dup_database.csv",
                         sep = "/"))
```

## 7.2 Save flags
Save the flags so far.
```{r 7.2}
BeeBDC::flagRecorder(
  data = check_time,
  outPath = paste(OutPath_Report, sep =""),
  fileName = paste0("flagsRecorded_", Sys.Date(),  ".csv"),
  idColumns = c("database_id", "id", "catalogNumber", "occurrenceID", "dataSource"),
  append = TRUE,
  printSummary = TRUE)
```

# 8.0 Data filtering 
The dataset can be re-read here if it does not already exist.
```{r 8.0}
if(!exists("check_time")){
 check_time <-
   readr::read_csv(paste(OutPath_Intermediate, "04_2_dup_database.csv",
                          sep = "/"))}
```

## 8.1 rm Outliers

Read in the most-recent duplicates file (generated by `dupeSummary()`) in order to identify the duplicates of the expert outliers.
```{r 8.1}
if(!exists("duplicates")){
  duplicates <- BeeBDC::fileFinder(path = DataPath,
                            fileName = "duplicateRun_") %>%
    readr::read_csv()}
```


Identify the outliers and get a list of their database_ids. This would require the source outlier files provided with the [BeeBDC](https://www.biorxiv.org/content/10.1101/2023.06.30.547152v1) paper. These files can further be modified to include more outliers.


    check_time <- BeeBDC::manualOutlierFindeR(
      data = check_time,
      DataPath = DataPath,
      PaigeOutliersName = "removedBecauseDeterminedOutlier.csv",
      newOutliersName = "^All_outliers_ANB_14March.xlsx",
      ColombiaOutliers_all = "All_Colombian_OutlierIDs.csv",
      # A .csv with manual outlier records that are too close to otherwise TRUE records
      NearTRUE = "nearTRUE.csv",
      duplicates = duplicates)



## 8.2 Save uncleaned 
Save the uncleaned dataset.

```{r 8.2}
  # Make sure that the .summary column is updated
check_time <- BeeBDC::summaryFun(
  data = check_time,
  dontFilterThese = c(".gridSummary", ".lonFlag", ".latFlag", ".uncer_terms",
                      ".uncertaintyThreshold"),
  removeFilterColumns = FALSE,
  filterClean = FALSE)
  # Save the uncleaned dataset
check_time %>% readr::write_excel_csv(.,
                                paste(OutPath_Intermediate, "05_unCleaned_database.csv",
                                      sep = "/"))
```


## 8.3 Filter 
Now clean the dataset of extra columns and failed rows and then save it.

```{r 8.3}
BeeBDC::summaryFun(
  data = check_time,
  dontFilterThese = c(".gridSummary", ".lonFlag", ".latFlag", ".uncer_terms",
                      ".uncertaintyThreshold"),
  # Remove the filtering columns?
  removeFilterColumns = TRUE,
  # Filter to ONLY cleaned data?
  filterClean = TRUE) %>% 
# Save this CLEANED dataset
  readr::write_excel_csv(.,
                   paste(OutPath_Intermediate, "05_cleaned_database.csv",
                         sep = "/"))
```

# 9.0 Figures and tables

## 9.1 Duplicate chordDiagrams 

Install BiocManager and ComplexHeatmap if you missed them at the start.
```{r 9.1, message=FALSE, warning=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")
```

Read in the most recent duplicate file, if it's not already in your environment.

```{r 9.1ii}
if(!exists("duplicates")){
  duplicates <- BeeBDC::fileFinder(path = DataPath,
                            fileName = "duplicateRun_") %>%
    readr::read_csv()}
```

Choose the global figure parameters.
```{r 9.1iii}
par(mar = c(2, 2, 2, 2)/2, mfrow = c(1,1))
```

Create the chorDiagram. You can leave many of the below values out but we show here the defaults. There are [internal] no duplicates in our test dataset and so BeeBDC will throw an informative error. However, we show the full output figure from our bee dataset below.

```{r 9.1iv, eval=FALSE, fig.fullwidth=TRUE, fig.height=7.5, fig.width=9}
BeeBDC::chordDiagramR(
  # The duplicate data from the dupeSummary function output  
  dupeData = duplicates,
  outPath = OutPath_Figures,
  fileName = "ChordDiagram.pdf",
  # These can be modified to help fit the final pdf that's exported.
  width = 9,
  height = 7.5,
  bg = "white",
  # How few distinct dataSources should a group have to be listed as "other"
  smallGrpThreshold = 3,
  title = "Duplicated record sources",
  # The default list of colour palettes to choose from usign the paleteer package
  palettes = c("cartography::blue.pal", "cartography::green.pal", 
               "cartography::sand.pal", "cartography::orange.pal", "cartography::red.pal",
               "cartography::purple.pal", "cartography::brown.pal"),
  canvas.ylim = c(-1.0,1.0), 
  canvas.xlim = c(-0.6, 0.25),
  text.col = "black",
  legendX = grid::unit(6, "mm"),
  legendY = grid::unit(18, "mm"),
  legendJustify = c("left", "bottom"),
  niceFacing = TRUE)
```

![Full chord diagram from Dorey et al. 2023]
<img src="https://github.com/jbdorey/BeeBDC/blob/main/inst/extdata/icon_vignettes/ChordDiagram.pdf" align="left" />


## 9.2 Duplicate histogram 

Read in the uncleaned dataset, if it's not already present.
```{r 9.2}
if(!exists("check_time")){
beeData <- readr::read_csv(paste(OutPath_Intermediate, "05_unCleaned_database.csv",
                                 sep = "/"),
                           col_types = BeeBDC::ColTypeR())
}else{
  beeData <- check_time
  rm(check_time)
}
```

Create a plot with two bar graphs. One shows the absolute number of duplicate records for each data source while the other shows the proportion of records that are duplicated within each data source. (*'dataSource'* is simplified to the text before the first underscore).
```{r 9.2ii, warning=FALSE}
BeeBDC::dupePlotR(
  data = beeData,
  # The outPath to save the plot as
  outPath = OutPath_Figures,
  fileName = "duplicatePlot.pdf",
  # Colours in order: duplicate, kept duplicate, unique
  dupeColours = c("#F2D2A2","#B9D6BC", "#349B90"),
  # Plot size and height
  base_height = 7, base_width = 7,
  legend.position = c(0.85, 0.8),
  # Extra variables can be fed into forcats::fct_recode() to change names on plot
  GBIF = "GBIF", SCAN = "SCAN", iDigBio = "iDigBio", USGS = "USGS", ALA = "ALA", 
  ASP = "ASP", CAES = "CAES", Ecd = "Ecd",
  returnPlot = TRUE
)
```

## 9.3 Flags by source 
Create a compound bar plot that shows the proportion of records that pass or fail each flag (rows) and for each data source (columns). The function can also optionally return a point map for a user-specified species when plotMap = TRUE. (*dataSource* is simplified to the text before the first underscore.)
```{r 9.3, fig.width=15, fig.height=9, fig.fullwidth=TRUE}
BeeBDC::plotFlagSummary(
  data = beeData,
  # Colours in order of pass (TRUE), fail (FALSE), and NA
  flagColours = c("#127852", "#A7002D", "#BDBABB"),
  fileName = paste0("FlagsPlot_", Sys.Date(),".pdf"),
  outPath = paste0(OutPath_Figures),
  width = 15, height = 9,
    # OPTIONAL:
      #   # Filter to a single species
      #       speciesName = "Holcopasites heliopsis",
      #         # column to look in
      #       nameColumn = "species",
      #        # Save the filtered data
      #       saveFiltered = TRUE,
      #   # Filter column to display on map
      #       filterColumn = ".summary",
      #       plotMap = TRUE,
      #   # amount to jitter points if desired, e.g. 0.25 or NULL
      #       jitterValue = NULL,
      #        # Map opacity value for points between 0 and 1
      #   mapAlpha = 1,
      #        # If a user wants to output the table used to make the figure, change this to TRUE
      #   saveTable = FALSE,
  # Extra variables can be fed into forcats::fct_recode() to change names on plot
  GBIF = "GBIF", SCAN = "SCAN", iDigBio = "iDigBio", USGS = "USGS", ALA = "ALA", 
  ASP = "ASP", CAES = "CAES", 'BMont' = "BMont", 'BMin' = "BMin", Ecd = "Ecd",
  Gaiarsa = "Gai", EPEL = "EPEL",
  returnPlot = TRUE
)
```

## 9.4 Maps 
Import CLEANED dataset (you can change this option).
```{r 9.4}
mapData <- readr::read_csv(paste(OutPath_Intermediate, "05_cleaned_database.csv",
                                 sep = "/"),
                           col_types = BeeBDC::ColTypeR())
```

### a. Summary maps
Draw a global summary map for occurrence and species number by country.
```{r 9.4a}
BeeBDC::summaryMaps(
  data = mapData,
  width = 10, height = 10,
  class_n = 3,
  class_Style = "fisher",
  fileName = "CountryMaps_fisher.pdf",
  outPath = OutPath_Figures
)
```

   
### b. Interactive maps
Use the occurrence data (preferably uncleaned in order to show pass/fail points) and outputs interactive .html maps, that can be opened in your browser, to a specific directory. The maps can highlight if an occurrence has passed all filtering (*.summary* == TRUE) or failed at least one filter (*.summary* == FALSE). This can be updated by first running `summaryFun()` to choose the columns that you want to be highlighted. It will also highlight occurrences flagged as expert-identified or country outliers separately. Because the function can have any categorical variable fed into 'speciesColumn', users may choose another column of interest to map; however, very large categories can be slow to produce and unwieldy to view.
```{r 9.4b}
BeeBDC::interactiveMapR(
   # occurrence data
  data = beeData,
   # Directory where to save files
  outPath = paste0(OutPath_Figures, "interactiveMaps", sep = "/"),
  lon = "decimalLongitude",
  lat = "decimalLatitude",
    # Occurrence dataset column with species names
  speciesColumn = "scientificName",
    # Which species to map — a character vector of names or "ALL"
    # Note: "ALL" is defined AFTER filtering for country
  speciesList = "ALL",
  countryList = NULL, # study area
    # Point jitter to see stacked points — jitters an amount in decimal degrees
  jitterValue = 0.01
)
```


## 9.5 Data providers
Read in the clean data if it's not already in the environment.
```{r 9.5}
if(!exists("mapData")){
  mapData <- readr::read_csv(paste(OutPath_Intermediate, "05_cleaned_database.csv",
                                sep = "/"),
                          col_types = BeeBDC::ColTypeR(),
  locale = readr::locale(encoding = "UTF-8"))}
```

This function will attempt to find and build a table of data providers that have contributed to the input data, especially using the *'institutionCode'* column. It will also look for a variety of other columns to find data providers using an internally set sequence of if-else statements. Hence, this function is quite specific for bee data, but should work for other taxa in similar institutions (perhaps to a lesser degree).
```{r 9.5ii}
dataProvTable <- BeeBDC::dataProvTables(data = mapData,
                                        runBeeDataChecks = TRUE,
                                        outPath = OutPath_Report,
                                        fileName = "dataProvTable.csv")
```


## 9.6 Flag summary
Takes a flagged dataset and returns the total number of fails (FALSE) per flag (columns starting with ".") and per species. Users may define the column to group the summary by. While it is intended to work with the *scientificName* column, users may select any grouping column (e.g., *country*).
```{r 9.6}
summaryTable <- BeeBDC::flagSummaryTable(data = beeData, 
                                         column = "scientificName", 
                                         outPath = OutPath_Report,
                                         fileName = "flagTable.csv")
```



Fin.


```{r 9.end, include=FALSE}
  # Secretely remove the RootPath once finished
unlink(paste0(RootPath), recursive = TRUE)
#unlink(paste0(here::here(), "/Output"), recursive = TRUE)
```

